---
title: "cdata"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The [`cdata`](https://github.com/WinVector/cdata) package is a demonstration of the ["coordinatized data" theory](http://winvector.github.io/FluidData/RowsAndColumns.html) and includes an implementation of the ["fluid data" methodology](http://winvector.github.io/FluidData/FluidData.html).   

Briefly `cdata` supplies data transform operators that:

 * Work on local data or with any `DBI` data source.
 * Are powerful generalizations of the operators commonly called `pivot` and `un-pivot`.
 * Can be specified by drawing an example.

A quick example:

```{r}
library("cdata")

# first few rows of the iris data as an example
d <- wrapr::build_frame(
   "Sepal.Length"  , "Sepal.Width", "Petal.Length", "Petal.Width", "Species" |
     5.1           , 3.5          , 1.4           , 0.2          , "setosa"  |
     4.9           , 3            , 1.4           , 0.2          , "setosa"  |
     4.7           , 3.2          , 1.3           , 0.2          , "setosa"  |
     4.6           , 3.1          , 1.5           , 0.2          , "setosa"  |
     5             , 3.6          , 1.4           , 0.2          , "setosa"  |
     5.4           , 3.9          , 1.7           , 0.4          , "setosa"  )
d$iris_id <- seq_len(nrow(d))

knitr::kable(d)
```

Now suppose we want to take the above "all facts about each iris are in a single row" representation and convert it into a per-iris record block with the following structure.

```{r}
record_example <- wrapr::qchar_frame(
   "plant_part"  , "measurement", "value"      |
     "sepal"     , "width"      , Sepal.Width  |
     "sepal"     , "length"     , Sepal.Length |
     "petal"     , "width"      , Petal.Width  |
     "petal"     , "length"     , Petal.Length )

knitr::kable(record_example)
```

The above sort of transformation may seem exotic, but it is fairly common when we want to plot many aspects of a record at the same time.

To specify our transformation we combine the record example with information about how records are keyed (recordKeys showing which rows go together to form a record, and controlTableKeys specifying the internal structure of a data record).

```{r}
layout <- rowrecs_to_blocks_spec(
  record_example,
  controlTableKeys = c("plant_part", "measurement"),
  recordKeys = c("iris_id", "Species"))

print(layout)
```

In the above we have used the common useful data organizing trick of specifying a dependent column (Species being a function of iris_id) as an additional key.

This layout then specifies and implements the data transform.  We can transform the data by sending it to the layout.

```{r}
d_transformed <- d %.>% 
  layout

knitr::kable(d_transformed)
```

And it is easy to invert these transforms using the `t()` transpose/adjoint notation.

```{r}
inv_layout <- t(layout)

print(inv_layout)

d_transformed %.>%
  inv_layout %.>%
  knitr::kable(.)
```

The layout specifications themselves are just simple lists with "pretty print methods" (the control table being simply and example record in the form of a data.frame).

```{r}
unclass(layout)
```

Notice that almost all of the time and space in using cdata is spent in specifying how your data is structured and is to be structured.

The current recommended interfaces to `cdata` are:

  * [`rowrecs_to_blocks_spec()`](https://winvector.github.io/cdata/reference/rowrecs_to_blocks_spec.html), for specifying how single row records relate to general multi-row (or block) records.
  * [`blocks_to_rowrecs_spec()`](https://winvector.github.io/cdata/reference/blocks_to_rowrecs_spec.html), for specifying how multi-row block records relate to single-row records.
  * [`layout_by()`](https://winvector.github.io/cdata/reference/layout_by.html) or the [wrapr dot arrow pipe](https://winvector.github.io/wrapr/reference/dot_arrow.html) for applying a layout to re-arrange data.
  * `t()` (transpose/adjoint) to invert or reverse layout specifications.
  * [`wrapr::qchar_frame()`](https://winvector.github.io/wrapr/reference/qchar_frame.html) a helper in specifying record control table layout specifications.
  * [`layout_specification()`](https://winvector.github.io/cdata/reference/layout_specification.html), for specifying transforms from multi-row records to other multi-row records.

The (older) recommended tutorial is: [Fluid data reshaping with cdata](http://winvector.github.io/FluidData/FluidDataReshapingWithCdata.html). We also have an (older) [short free cdata screencast](https://youtu.be/4cYbP3kbc0k) (and another example can be found [here](http://winvector.github.io/FluidData/DataWranglingAtScale.html)).

