---
title: "Control Table Keys"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Control Table Keys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In our [`cdata`](https://github.com/WinVector/cdata) [`R`](https://www.r-project.org) package and training materials we emphasize the [record-oriented thinking](https://winvector.github.io/cdata/articles/blocksrecs.html) and [how to design a transform control table](https://winvector.github.io/cdata/articles/design.html).  We now have an additional exciting new feature: control table keys.

The user can now control which columns of a `cdata` control table are the keys, including now using composite keys (that is keys that are spread across more than one column).  This is easiest to demonstrate with an example.

Consider the simple data frame (the first couple of rows of the famous `iris` data set).

```{r}
library("cdata")

d <- wrapr::build_frame(
   "Sepal.Length"  , "Sepal.Width", "Petal.Length", "Petal.Width", "Species" |
     5.1           , 3.5          , 1.4           , 0.2          , "setosa"  |
     4.9           , 3            , 1.4           , 0.2          , "setosa"  )
d$id <- seq_len(nrow(d))

knitr::kable(d)
```

Suppose we wish to land the dimensions of different irises in rows keyed by two columns: `Part` and `Measure`. That is we want the data to look like the following.

```{r}
expect <- wrapr::build_frame(
   "id"  , "Species", "Part" , "Measure", "Value" |
     1L  , "setosa" , "Sepal", "Length" , 5.1     |
     1L  , "setosa" , "Sepal", "Width"  , 3.5     |
     1L  , "setosa" , "Petal", "Length" , 1.4     |
     1L  , "setosa" , "Petal", "Width"  , 0.2     |
     2L  , "setosa" , "Sepal", "Length" , 4.9     |
     2L  , "setosa" , "Sepal", "Width"  , 3       |
     2L  , "setosa" , "Petal", "Length" , 1.4     |
     2L  , "setosa" , "Petal", "Width"  , 0.2     )

knitr::kable(expect)
```

This may seem like an exotic transform, but it seems to be common when you want to use multiple display aesthetics in a plot (such as x, y, color, and facet).

With multiple control table keys specifying these sorts of transforms can be made easy.

First define the control table specifying what a single row-record looks like after transformation to a multi-row block-record.

```{r}
control_table <- wrapr::qchar_frame(
   "Part"   , "Measure", "Value"      |
     "Sepal", "Length" , Sepal.Length |
     "Sepal", "Width"  , Sepal.Width  |
     "Petal", "Length" , Petal.Length |
     "Petal", "Width"  , Petal.Width  )

layout <- rowrecs_to_blocks_spec(
  control_table,
  controlTableKeys = c("Part", "Measure"),
  recordKeys = c("id", "Species"))

print(layout)
```

Notice we are specifying that the columns `Part` and `Measure` are the control table row-keys (identifying rows within the block record) and "id" and "Species" are record keys (specifying which rows are in the same record).

We have designed our desired transform above (as taught [here](https://winvector.github.io/cdata/articles/design.html)). We are introducing a new convention of putting the key values in quotes (even though `wrapr::qcar_frame()` does not need this).  Notice the quoted fields are column-names of the original row-oriented data, and the quote is indicating that these are names standing in for values.

We can now perform the transform.  

```{r}
res <- d %.>% layout

knitr::kable(res)
```

And, as one comes to expect with coordinatized/fluid-data transforms, the process is easy to reverse (modulo column and row order).

```{r}
inv_layout <- t(layout)

print(inv_layout)

back <- res %.>% inv_layout

knitr::kable(back)
```

Recle Etino Vibal ([who asked for this feature in an issue](https://github.com/WinVector/cdata/issues/3)) has [an interesting article](https://amateurdatasci.rbind.io/post/table-another-back-again-cdata/) trying some variations on the data shaping concepts.

The `cdata` unit test include the following variations of the above example:

  * [A simple "there and back example"](https://github.com/WinVector/cdata/blob/master/inst/unit_tests/test_there_and_back.R).
  * [The "there and back" example on a database](https://github.com/WinVector/cdata/blob/master/inst/unit_tests/test_there_and_back_db.R).
  * [A composite control keys example](https://github.com/WinVector/cdata/blob/master/inst/unit_tests/test_composite_control_keys.R) (this demo).
  * [The composite control keys example on a database](https://github.com/WinVector/cdata/blob/master/inst/unit_tests/test_composite_control_keys_db.R).
  
We think `cdata` (and the accompanying [fluid data methodology](https://winvector.github.io/FluidData/FluidData.html), plus [extensions](https://winvector.github.io/cdata/articles/general_transform.html)) is a very deep and powerful way of wrangling data.  Once you take the time to learn the methodology (which is "draw what you want to happen to one record, type that in as your control table, and you are done!") it is very easy to use.


  