---
title: "Operators"
author: "John Mount, Nina Zumel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

As of `cdata` version `1.0.8` `cdata` implements an operator notation for data transform.

The idea is simple, yet powerful.

First let's start with some data.

```{r}

d <- wrapr::build_frame(
  "id", "measure", "value" |
    1   , "AUC"    , 0.7     |
    1   , "R2"     , 0.4     |
    2   , "AUC"    , 0.8     |
    2   , "R2"     , 0.5     )

knitr::kable(d)
```

In the above data we have two measurements each for two individuals (individuals identified by the "`id`" column).  Using `cdata`'s `new_record_spec()` method we can capture a description of this record structure.

```{r}
library("cdata")

record_spec <- new_record_spec(
  wrapr::build_frame(
    "measure", "value" |
    "AUC"    , "AUC" |
    "R2"     , "R2"  ),
  recordKeys = "id")

print(record_spec)
```

Once we have this specification we can transform the data using operator notation.

We can collect the record blocks into rows by a "division" (or aggregation/projection) step.

```{r}
knitr::kable(d)

d2 <- d %//% record_spec

knitr::kable(d2)
```

We can expand record rows into blocks by a "multiplication" (or join) step.

```{r}
knitr::kable(d2)

d3 <- d2 %**% record_spec

knitr::kable(d3)
```

And that is trully fluid data manipulation.

